---
title: "Visualizing Inequalities in Life Expectancy"
author: "Xiling Zhu"
date: "3/31/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
setwd("~/data-camp-guided-projects/Visualizing Inequalities in Life Expectancy")
library(dplyr)
library(tidyr)
library(ggplot2)

# This sets plot images to a nice size
options(repr.plot.width = 6, repr.plot.height = 6)
```


## United Nations life expectancy data
Life expectancy at birth is a measure of the average a living being is expected to live. It takes into account several demographic factors like gender, country, or year of birth.

Life expectancy at birth can vary along time or between countries because of many causes: the evolution of medicine, the degree of development of countries, or the effect of armed conflicts. Life expectancy varies between gender, as well. The data shows that women live longer that men. Why? Several potential factors, including biological reasons and the theory that women tend to be more health conscious.

Let's create some plots to explore the inequalities about life expectancy at birth around the world. We will use a dataset from the United Nations Statistics Division, which is available here(http://data.un.org/Data.aspx?d=GenderStat&f=inID:37&c=1,2,3,4,5,6&s=crEngName:asc,sgvEngName:asc,timeEngName:desc&v=1).

```{r loading data, echo=FALSE}
life_expectancy <- read.csv("datasets/UNdata.csv")
head(life_expectancy)
```

## Life expectancy of men vs. women by country

Let's manipulate the data to make our exploration easier. We will build the dataset for our first plot in which we will represent the average life expectancy of men and women across countries for the last period recorded in our data (2000-2005).

```{r average life expectancy of men and women across countries 2000-2005, echo=FALSE}
# Reshaping the long data into wide data
subdata <- life_expectancy  %>% 
    filter(Year == "2000-2005")%>%
    select(Country.or.Area, Subgroup, Value)  %>% 
    spread(Subgroup, Value)

head(subdata)
```

## Visualize average life expectancy of men and women across countries from 2000 to 2005
```{r Visualize average life expectancy of men and women across countries 2000-2005, echo=FALSE}
# Plotting male and female life expectancy
sub_plot <- ggplot(subdata, aes(x = Male, y = Female))+
            geom_point()+
            geom_abline(intercept = 0, slope = 1, linetype = 2)+
            scale_x_continuous(limits = c(35, 85))+
            scale_y_continuous(limits = c(35, 85))+
            labs(title="Life Expectancy at Birth by Country",
                 subtitle="Years. Period: 2000-2005. Average.",
                 caption="Source: United Nations Statistics Division",
                 x="Males",
                 y="Females")
sub_plot
```

## Highlighting countries where gender gap in average life expectancy from 2000 to 2005 is high
```{r Highlighting countries}
# Subsetting top 3 countries where gender gap in life expectancy is high
top_male <- subdata %>% arrange(Male-Female) %>% head(3)
top_female <- subdata %>% arrange(Female-Male) %>% head(3)

# Adding text to the previous plot to label countries of interest
ggplot(subdata, aes(x=Male, y=Female, label = Country.or.Area))+
  geom_point(colour="white", fill="chartreuse3", shape=21, alpha=.55, size=5)+
  geom_abline(intercept = 0, slope = 1, linetype=2)+
  scale_x_continuous(limits=c(35,85))+
  scale_y_continuous(limits=c(35,85))+
  labs(title="Life Expectancy at Birth by Country",
       subtitle="Years. Period: 2000-2005. Average.",
       caption="Source: United Nations Statistics Division",
       x="Males",
       y="Females")+
    geom_text(data = top_male, size = 3)+
  geom_text(data = top_female, size = 3)+
  theme_bw()
```
## How has life expectancy by gender envolved?
Since our data contains historical information, let's see now how life expectancy has evolved in recent years. Our second plot will represent the difference between men and women across countries between two periods: 2000-2005 and 1985-1990.

Let's start building a dataset called subdata2 for our second plot.
```{r Build subdata2 evolution of life expectancy by gender, echo=FALSE}
# Subsetting, mutating and reshaping the life expectancy data
subdata2 <- life_expectancy %>% 
  filter(Year %in% c("1985-1990", "2000-2005")) %>% 
  mutate(Sub_Year=paste(Subgroup, Year, sep="_")) %>% 
  mutate(Sub_Year=gsub("-", "_", Sub_Year)) %>% 
  select(-Subgroup, -Year) %>%  
  spread(Sub_Year, Value)  %>% 
  mutate(diff_Female = Female_2000_2005 - Female_1985_1990)%>%
  mutate(diff_Male = Male_2000_2005 - Male_1985_1990)

# Taking a look at the first few rows
head(subdata2)
```
## Visualize how has life expectancy by gender has evolved
Now let's create our second plot in which we will represent average life expectancy differences between "1985-1990" and "2000-2005" for men and women.
```{r Visualize the evolution of life expectancy by gender}
ggplot(subdata2, aes(x=diff_Male, y=diff_Female, label=Country.or.Area))+
  geom_point(colour="white", fill="chartreuse3", shape=21, alpha=.55, size=5)+
  geom_abline(intercept = 0, slope = 1, linetype=2)+
  scale_x_continuous(limits=c(-25,25))+
  scale_y_continuous(limits=c(-25,25))+
  geom_hline(yintercept = 0, linetype = 2)+
  geom_vline(xintercept = 0, linetype = 2)+
  labs(title="Life Expectancy at Birth by Country",
       subtitle="Years. Difference between 1985-1990 and 2000-2005. Average.",
       caption="Source: United Nations Statistics Division",
       x="Males",
       y="Females")+
theme_bw()
```
10. Highlighting remarkable countries in the evolution of gender gap of life expectancy
As we did in the first plot, let's label some points. Concretely, we will point those three where the aggregated average life expectancy for men and women increased most and those three where decreased most in the period.
```{r Highlighting remarkable countries}
top <- subdata2 %>% arrange(diff_Male+diff_Female) %>% head(3)
bottom <- subdata2 %>% arrange(desc(diff_Male+diff_Female)) %>% head(3)

# Adding text to the previous plot to label countries of interest
ggplot(subdata2, aes(x=diff_Male, y=diff_Female, label=Country.or.Area), guide=FALSE)+
  geom_point(colour="white", fill="chartreuse3", shape=21, alpha=.55, size=5)+
  geom_abline(intercept = 0, slope = 1, linetype=2)+
  scale_x_continuous(limits=c(-25,25))+
  scale_y_continuous(limits=c(-25,25))+
  geom_hline(yintercept=0, linetype=2)+
  geom_vline(xintercept=0, linetype=2)+
  labs(title="Life Expectancy at Birth by Country",
       subtitle="Years. Difference between 1985-1990 and 2000-2005. Average.",
       caption="Source: United Nations Statistics Division",
       x="Males",
       y="Females")+
  geom_text(data = top)+
  geom_text(data = bottom)+
  theme_bw()
```

