---
title: "Predict Taxi Fares with Random Forests"
author: "Xiling Zhu"
date: "4/2/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## 1. 49999 New York taxi trips

To drive a yellow New York taxi, you have to hold a "medallion" from the city's Taxi and Limousine Commission. Recently, one of those changed hands for over one million dollars, which shows how lucrative the job can be.

But this is the age of business intelligence and analytics! Even taxi drivers can stand to benefit from some careful investigation of the data, guiding them to maximize their profits. In this project, we will analyze a random sample of 49999 New York journeys made in 2013. We will also use regression trees and random forests to build a model that can predict the locations and times when the biggest fares can be earned.

Let's start by taking a look at the data!

```{r Load package and datasets, warning=FALSE, error=FALSE, message=FALSE}
# Loading the tidyverse
library(tidyverse)

# Reading in the taxi data
taxi <- read_csv("datasets/taxi.csv")

# Taking a look at the first few rows in taxi
head(taxi)
```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
